---
title: "Propensity Score Matching Example"
author: "Data Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#check that packages are installed, if not install them
if (!requireNamespace("MatchIt", quietly = TRUE)) {
  install.packages("MatchIt")
}

# check if cobalt is installed
if (!requireNamespace("cobalt", quietly = TRUE)) {
  install.packages("cobalt")
}


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(MatchIt)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(cobalt)


```

## Introduction

This example demonstrates propensity score matching using simulated data. We'll show histograms of covariate distributions before and after matching to assess balance improvement.

## Generate Sample Data

```{r generate_data}
set.seed(123)
n <- 1000

# Generate covariates
age <- rnorm(n, 50, 15)
income <- rnorm(n, 40000, 15000)
education <- sample(1:5, n, replace = TRUE)
health_score <- rnorm(n, 70, 20)

# Generate treatment assignment (with selection bias)
treatment_prob <- plogis(-2 + 0.02*age + 0.00003*income + 0.3*education + 0.01*health_score)
treatment <- rbinom(n, 1, treatment_prob)

# Generate outcome with treatment effect
outcome <- 10 + 2*treatment + 0.1*age + 0.0001*income + 1.5*education + 0.05*health_score + rnorm(n, 0, 5)

# Create dataset
data <- data.frame(
  treatment = treatment,
  age = age,
  income = income,
  education = education,
  health_score = health_score,
  outcome = outcome
)

print(paste("Treatment group size:", sum(data$treatment)))
print(paste("Control group size:", sum(1-data$treatment)))
```

## Pre-matching Analysis

```{r pre_matching_balance}
# Check balance before matching
bal_before <- bal.tab(treatment ~ age + income + education + health_score, 
                      data = data, estimand = "ATT")
print("Balance before matching:")
print(bal_before)
```

## Histograms Before Matching

```{r histograms_before, fig.width=12, fig.height=8}
# Create histograms for each covariate before matching
p1_before <- ggplot(data, aes(x = age, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Age Distribution Before Matching", x = "Age", y = "Count", fill = "Group") +
  theme_minimal()

p2_before <- ggplot(data, aes(x = income, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Income Distribution Before Matching", x = "Income", y = "Count", fill = "Group") +
  theme_minimal()

p3_before <- ggplot(data, aes(x = education, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 5) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Education Distribution Before Matching", x = "Education Level", y = "Count", fill = "Group") +
  theme_minimal()

p4_before <- ggplot(data, aes(x = health_score, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Health Score Distribution Before Matching", x = "Health Score", y = "Count", fill = "Group") +
  theme_minimal()

grid.arrange(p1_before, p2_before, p3_before, p4_before, ncol = 2)
```

## Propensity Score Matching

```{r matching}
# Perform propensity score matching
match_out <- matchit(treatment ~ age + income + education + health_score,
                     data = data,
                     method = "nearest",
                     distance = "glm",
                     caliper = 0.1)

# Summary of matching
summary(match_out)

# Extract matched data
matched_data <- match.data(match_out)
```

## Post-matching Balance

```{r post_matching_balance}
# Check balance after matching
bal_after <- bal.tab(match_out, estimand = "ATT")
print("Balance after matching:")
print(bal_after)
```

## Histograms After Matching

```{r histograms_after, fig.width=12, fig.height=8}
# Create histograms for each covariate after matching
p1_after <- ggplot(matched_data, aes(x = age, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Age Distribution After Matching", x = "Age", y = "Count", fill = "Group") +
  theme_minimal()

p2_after <- ggplot(matched_data, aes(x = income, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Income Distribution After Matching", x = "Income", y = "Count", fill = "Group") +
  theme_minimal()

p3_after <- ggplot(matched_data, aes(x = education, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 5) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Education Distribution After Matching", x = "Education Level", y = "Count", fill = "Group") +
  theme_minimal()

p4_after <- ggplot(matched_data, aes(x = health_score, fill = factor(treatment))) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Health Score Distribution After Matching", x = "Health Score", y = "Count", fill = "Group") +
  theme_minimal()

grid.arrange(p1_after, p2_after, p3_after, p4_after, ncol = 2)
```

## Before vs After Comparison

```{r comparison_plots, fig.width=15, fig.height=10}
# Side-by-side comparison for age
grid.arrange(p1_before, p1_after, 
             p2_before, p2_after,
             p3_before, p3_after,
             p4_before, p4_after,
             ncol = 2,
             top = "Covariate Distributions: Before (Left) vs After (Right) Matching")
```

## Treatment Effect Estimation

```{r treatment_effect}
# Estimate treatment effect before matching
effect_before <- lm(outcome ~ treatment + age + income + education + health_score, data = data)
print("Treatment effect before matching:")
summary(effect_before)$coefficients[2, ]

# Estimate treatment effect after matching
effect_after <- lm(outcome ~ treatment + age + income + education + health_score, 
                   data = matched_data, weights = matched_data$weights)
print("Treatment effect after matching:")
summary(effect_after)$coefficients[2, ]

# Simple difference in means after matching
mean_diff <- with(matched_data, mean(outcome[treatment == 1]) - mean(outcome[treatment == 0]))
print(paste("Simple mean difference after matching:", round(mean_diff, 3)))
```

## Propensity Score Distribution

```{r propensity_scores, fig.width=10, fig.height=6}
# Plot propensity score distributions
data$prop_score <- match_out$distance
matched_data$prop_score <- matched_data$distance

p_prop_before <- ggplot(data, aes(x = prop_score, fill = factor(treatment))) +
  geom_histogram(alpha = 0.3, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Propensity Score Before Matching", x = "Propensity Score", y = "Count", fill = "Group") +
  theme_minimal()

p_prop_after <- ggplot(matched_data, aes(x = prop_score, fill = factor(treatment))) +
  geom_histogram(alpha = 0.3, position = "identity", bins = 30) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), 
                    labels = c("Control", "Treatment")) +
  labs(title = "Propensity Score After Matching", x = "Propensity Score", y = "Count", fill = "Group") +
  theme_minimal()

grid.arrange(p_prop_before, p_prop_after, ncol = 2)
```

## Conclusion

This example demonstrates how propensity score matching can improve covariate balance between treatment and control groups. The histograms show that after matching:

1. The distributions of covariates between treatment and control groups become more similar
2. The overlap in propensity scores improves
3. The standardized mean differences decrease, indicating better balance

The matched dataset provides a more reliable basis for estimating causal treatment effects by reducing selection bias.