---
title: "Analysis of Heart Disease"
output:
  html_notebook: default
  pdf_document: default
  word_document: default
---

# Libraries
```{r}
library(tidyverse)
library(here)
```


```{r}
# get current working directory
getwd()


# Set working directory to "/Users/williamparker/Desktop/PSOMERCAMP_databootcamp"
setwd("/Users/williamparker/Desktop/PSOMERCAMP_databootcamp")

```

# Load Data
```{r}
# Load the pre-processed heart disease dataset
heart_data <- read_csv("/Users/williamparker/Desktop/PSOMERCAMP_databootcamp/data/heart_data.csv")
```

# Results
```{r}
# construct binary primary outcome of any_heart_disease from num > 0
heart_data <- heart_data %>%
  mutate(any_heart_disease = ifelse(num > 0, 1, 0))

proportion_heart_disease <- mean(heart_data$any_heart_disease, na.rm = TRUE)
```

## Study population
There are `r nrow(heart_data)` observations in the dataset, with `r sum(heart_data$any_heart_disease)` cases of heart disease. The proportion of individuals with heart disease is `r round(proportion_heart_disease * 100, 2)`%. The mean age is `r round(mean(heart_data$age, na.rm = TRUE), 1)` years, and the mean cholesterol level is `r round(mean(heart_data$chol, na.rm = TRUE), 1)` mg/dl and the proportion of men is `r round(mean(heart_data$sex, na.rm = TRUE) * 100, 1)`%.

# Logistic Regression Model
This section performs a logistic regression to model the likelihood of heart disease based on age and cholesterol levels.

```{r}
# Fit the logistic regression model
# The outcome variable is any_heart_disease
# The predictor variables are age and chol
logistic_model <- glm(any_heart_disease ~ age + chol, 
                      data = heart_data, 
                      family = binomial)

# Display the summary of the model to see the results
summary(logistic_model)

# Load required libraries
library(broom)
library(dplyr)
library(flextable)
library(officer)


# Tidy the regression output
tidy_model <- tidy(logistic_model)

# Optionally round for nicer display
tidy_model <- tidy_model %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

# Create a flextable
ft <- flextable(tidy_model)
ft <- autofit(ft)
ft <- set_caption(ft, caption = "Logistic Regression Results")

# Export to Word document
doc <- read_docx() %>%
  body_add_par("Logistic Regression Summary", style = "heading 1") %>%
  body_add_flextable(ft)

# Save the document
print(doc, target = "logistic_regression_summary.docx")

```

# Cholesterol vs. Age by Heart Disease Status
```{r}
# Create a scatter plot of cholesterol vs. age, colored by heart disease status
heart_data %>% 
    mutate(any_heart_disease = factor(any_heart_disease, levels = c(0,1), labels = c("No", "Yes"))) %>% 
    ggplot(aes(x = age, y = chol, color = any_heart_disease)) + 
    geom_point(alpha=0.7) + 
    labs(title = "Cholesterol vs. Age by Heart Disease Status",
        x = "Age",
        y = "Cholesterol (mg/dl)",
        color = "Heart Disease") + 
    theme_minimal()
```


