---
title: "Data set prep for PSOMER CAMP project"
output: html_notebook
---

# Libraries

It's important you know what each package is doing
```{r}
# for data manipulation and visualization
library(tidyverse)

# to help set the working directory and make this script reproducible across different machines
library(here)

# for Table one
library(tableone) 
```

## check the working directory- this should be the filepath to your project folder

```{r}
here::here()
```

# Methods

## Study population and inclusion criteria

This study identifies all....

### Load in dataset

Load in your dataset for your project. For this example, I am using an example dataset from: 

https://archive.ics.uci.edu/dataset/45/heart+disease

```{r}
# you are unlikely to need col_names = FALSE
heart_data <- read_csv(here("data", "processed.cleveland.data"), col_names = FALSE)

# here I am labeling the columns- you might need to do this!
column_names <- c(
  "age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
  "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num"
)

heart_data <- heart_data %>%
  setNames(column_names)
```

## Primary outcome

The primary outcome of this study is the diagnosis of "any heart disease", defined as....
```{r}
# construct binary primary outcome of any_heart_disease from num > 0
heart_data <- heart_data %>%
  mutate(any_heart_disease = ifelse(num > 0, 1, 0))
```

# Exploratory data analysis

This section won't go in your methods, but is essential to understand your dataset

### view your dataset
```{r}
heart_data
```

# Create a table one
```{r}
# Create a table one for the final cohort
table_one <- CreateTableOne(data = heart_data)

table_one
```



## Exploratory data analysis

Visualize distribution of age in the dataset
```{r}
heart_data %>%
  ggplot(aes(x = age)) +
  geom_density(fill = "blue", alpha = 0.5)
```

## calculate the mean, median, and mode of age

```{r}
mean_age <- mean(heart_data$age, na.rm = TRUE)
median_age <- median(heart_data$age, na.rm = TRUE)
mode_age <- as.numeric(names(sort(table(heart_data$age), decreasing = TRUE)[1]))
```

```{r}
mean_age
```
```{r}
median_age
```

```{r}
mode_age
```

```{r}
heart_data %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1,
                 fill = "blue", color = "black", alpha = 0.5)
```


```{r}
heart_data %>%
  group_by(any_heart_disease) %>%
  ggplot(aes(x = age, fill = factor(any_heart_disease))) +
  geom_density(alpha = 0.5)
```


## test the hypothesis that patients with heart disease are older on average

specifically test that the mean age is higher in patients with heart disease compared to those without


```{r}
mean_age_heart_disease <- heart_data %>%
  filter(any_heart_disease == 1) %>%
  summarise(mean_age = mean(age, na.rm = TRUE))

mean_age_heart_disease
```
### mean age of patients without heart disease
```{r}
mean_age_no_heart_disease <- heart_data %>%
  filter(any_heart_disease == 0) %>%
  summarise(mean_age = mean(age, na.rm = TRUE))

mean_age_no_heart_disease
```
### null hypothesis

the mean age is the same in patients with and without heart disease

mean_age_heart_disease == mean_age_no_heart_disease

### formula for a t-test in latex

$$t = \frac{\bar{X}_1 - \bar{X}_2}{s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$


$$s_p = \sqrt{\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}}$$

```{r}
t_test_result <- t.test(age ~ any_heart_disease, data = heart_data)

t_test_result
```
## Unadjusted association of age with heart disease

```{r}
# Create a logistic regression model to assess the association of age with heart disease
logistic_model <- glm(any_heart_disease ~ age, data = heart_data, family = binomial)
```

```{r}
summary(logistic_model)
```
```{r}
exp(0.05199)
```

## scatterplot of age vs cholesterol
```{r}
heart_data %>%
  ggplot(aes(x = age, y = chol)) +
  geom_point() +
  labs(title = "Scatterplot of Age vs Cholesterol",
       x = "Age",
       y = "Cholesterol")
```

```{r}
heart_data %>%
  mutate(any_heart_disease = factor(any_heart_disease, levels = c(0, 1), labels = c("No", "Yes"))) %>%
  ggplot(aes(x= age, y = thalach, color = factor(any_heart_disease))) + geom_point() + labs(x = "age", y = "maximum heart rate", color = "heart disease")
```


### adujsted logistic regression model controlling for cholesterol
```{r}
logistic_model_adjusted <- glm(any_heart_disease ~ age + chol, 
                               data = heart_data, family = binomial)
```

```{r}
summary(logistic_model_adjusted)
```







