---
title: "Data set prep for PSOMER CAMP project"
output: html_notebook
---

# Libraries

It's important you know what each package is doing
```{r}
# for data manipulation and visualization
library(tidyverse)

# to help set the working directory and make this script reproducible across different machines
library(here)

# for the strobe diagram
library(DiagrammeR)
library(DiagrammeRsvg) 
library(glue)
library(rsvg)

# for Table one
library(tableone) 
```

## check the working directory- this should be the filepath to your project folder
```{r}
here::here()
```


# Methods

## Study population and inclusion criteria

This study identifies all....

### Load in dataset
Load in your dataset for your project. For this example, I am simulating a data set
```{r}
# uncomment the line below
# data <- read_csv(here("data", "datafile.csv"))

## simulate a dataset. There are 1500 patients. The outcome is binary, and there are 5 covariates (age, creatinine, diabetes, BMI, and smoking status)
set.seed(123)
data <- tibble(
  patient_id = 1:1500,
  age = rnorm(1500, mean = 50, sd = 10),
  creatinine = rnorm(1500, mean = 1.0, sd = 0.2),
  diabetes = sample(c(0, 1), 1500, replace = TRUE),
  BMI = rnorm(1500, mean = 25, sd = 5),
  smoking_status = sample(c("non-smoker", "smoker"), 1500, replace = TRUE)
)

# create a noisy binary outcome variable
data <- data %>%
  mutate(risk = age + creatinine + diabetes * 2 + BMI - ifelse(smoking_status == "smoker", 1, 0),
         # add some noise to the risk score
         risk = risk + rnorm(1500, mean = 0, sd = 5),
         # create a binary outcome based on the risk score
         outcome = ifelse(risk > quantile(risk, 0.48), 1, 0))

data %>%
  count(outcome)
```

### Select only the relevant variables for your study
```{r}
data <- data %>% 
  select(patient_id, age, creatinine, diabetes, BMI, smoking_status, outcome)

data
```

### Define inclusion criteria and filter the dataset accordingly
Filter the dataset to include only patients who meet the inclusion criteria. For this example, let's say we want to include patients aged 30-70, with a creatinine level between 0.5 and 1.5.

Track of the patients you exclude for your STROBE diagram
```{r}
# Define inclusion criteria
data <- data %>%
  mutate(inclusion = ifelse(age >= 30 & age <= 70 & creatinine >= 0.5 & creatinine <= 1.5, TRUE, FALSE)) 

data %>%
  count(inclusion)
```

### Exclude patients who do not meet the inclusion criteria and save them in dataset
```{r}
excluded_patients <- data %>%
  filter(inclusion == FALSE) %>%
  select(patient_id, age, creatinine, diabetes, BMI, smoking_status, outcome)
```


### Make a strobe flow diagram with the exclusion reasons

Use generative AI for this! Do NOT code it by hand
```{r}
# Starting cohort
total_n <- nrow(data)

# Exclusion: Age < 30
excl_age_low <- data %>% filter(age < 30)
n_excl_age_low <- nrow(excl_age_low)
data <- data %>% filter(age >= 30)

# Exclusion: Age > 70
excl_age_high <- data %>% filter(age > 70)
n_excl_age_high <- nrow(excl_age_high)
data <- data %>% filter(age <= 70)

# Exclusion: Creatinine < 0.5
excl_cr_low <- data %>% filter(creatinine < 0.5)
n_excl_cr_low <- nrow(excl_cr_low)
data <- data %>% filter(creatinine >= 0.5)

# Exclusion: Creatinine > 1.5
excl_cr_high <- data %>% filter(creatinine > 1.5)
n_excl_cr_high <- nrow(excl_cr_high)
data <- data %>% filter(creatinine <= 1.5)

# Final cohort
final_n <- nrow(data)
n_excluded <- total_n - final_n

#save final cohort
write_csv(data, here("data", "final_cohort.csv"))
```

```{r}
strobe_diagram <- grViz(glue("
digraph flowchart {{
  rankdir = TB;
  node [shape = box, style = filled, fontname = Helvetica]
  edge [fontname = Helvetica]

  A [label = 'Initial cohort\\n(n = {total_n})', fillcolor = lightgray]
    B [label = 'Excluded (n = {n_excluded})\\n{n_excl_age_low} Age < 30\\n{n_excl_age_high} Age > 70\\n{n_excl_cr_low} Creatinine < 0.5\\n{n_excl_cr_high} Creatinine > 1.5', fillcolor = lightblue]
  C [label = 'Final cohort\\n(n = {final_n})', fillcolor = lightgray]
  X [label = '', width=0, height=0, shape=point, style=invis]

  A -> X [arrowhead=none]
  X -> B
  X -> C 

  {{ rank = same; B; X }}
}}
"))

strobe_diagram
```


```{r}
#convert the DiagrammeR object to SVG
svg <- export_svg(strobe_diagram)

# Convert SVG to PDF and save it
rsvg_pdf(charToRaw(svg), here("results", "strobe_diagram.pdf"))
```




# Create a table one
```{r}
# Create a table one for the final cohort
table_one <- CreateTableOne(vars = c("age", "creatinine", "diabetes", "BMI", "smoking_status", "outcome"),
                             data = data,
                             factorVars = c("diabetes", "smoking_status", "outcome"))

table_one


```
```{r}
#export to results
# Convert to data frame
table_df <- as.data.frame(print(table_one, quote = FALSE, noSpaces = TRUE))

# Write CSV
write.csv(table_df, file = here("results", "table_one.csv"), row.names = TRUE)
```

